---
title: "PA1_template"
author: "Fides"
date: "2/25/2021"
output: html_document
---

# 1-Loading and preprocessing the data
### a) Load the data and packages
```{r}
activity <- read.csv(file="~/Desktop/activity.csv", header=TRUE, sep=",", na.strings = "NA")

library("dplyr")
library("ggplot2")
library("data.table")
library("lubridate")
library("httpuv")
```
### b) Pre-process the data
##Unify date class
```{r}
activity$date <- ymd(activity$date)
```
##Remove missing values
```{r}
activity1 <- na.omit(activity)
```
##Create new variables 
```{r}
activity$monthly <- as.numeric(format(activity$date, "%m"))
activity$daily <- as.numeric(format(activity$date, "%d"))
```
##Check data
```{r}
summary(activity1)
```
## 2-Calculate total number of steps
### a) Make summary of data for ggplot
```{r}
activity2 <- summarize(group_by(activity1, date), daily.step=sum(steps))
mean.activity <- as.integer(mean(activity2$daily.step))
median.activity <- as.integer(median(activity2$daily.step))
```
### b) Make a histogram of the total number of steps taken each day
```{r}
plot.steps.day <- ggplot(activity2, aes(x=daily.step)) + 
  geom_histogram(binwidth = 1000, aes(y=..count.., fill=..count..)) + 
  geom_vline(xintercept=mean.activity, colour="orange", linetype="dotted", size=1) +
  geom_vline(xintercept=median.activity, colour="blue" , linetype="dashed", size=1) +
  labs(title="Number of Steps taken each day (histogram)", y="Frequency", x="Steps each day") 
plot.steps.day
```
### c) Mean total number of steps taken per day
```{r}
mean.activity
```
### c) Median total number of steps taken per day
```{r}
median.activity
```
## 3-What is the average daily activity pattern?
##Prepare data for ggplot
```{r}
activity3 <- activity1 %>% 
        group_by(interval) %>% 
        summarize(mean.step=mean(steps))
```
### a) Make a time series plot (i.e. \color{red}{\verb|type = "l"|}type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
plot.step.interval <- ggplot(activity3, aes(x=interval,y=mean.step)) + 
  geom_line(color="blue") + 
  labs(title="Average Number of Steps Taken - 5-min Interval", y="Average Number of Steps", x="5-min Intervals")
plot.step.interval
```

### b) Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
optimal <- which.max(activity3$mean.step)
optimal.step <- activity3$interval[optimal]
sprintf("%gth 5-min interval has maximum number of steps", optimal.step)
```
## 4-Imputing missing values
### a) Total number of missing values in dataset
```{r}
sum(is.na(activity))
```
### b) Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
```{r}
impute.activity <- activity
impute.activity$steps[is.na(impute.activity$steps)] <- mean(impute.activity$steps,na.rm=TRUE)
impute.activity$steps <- as.numeric(impute.activity$steps)
impute.activity$interval <- as.numeric(impute.activity$interval)
colSums(is.na(impute.activity))
```
### c) Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
summary(impute.activity)
```
##Summarize data by date
```{r}
impute.activity2 <- summarize(group_by(impute.activity,date),daily.step=sum(steps))

mean.impute   <- as.integer(mean(impute.activity2$daily.step))
mean.impute
median.impute <- as.integer(median(impute.activity2$daily.step))
median.impute
```
### d) Plot histogram
```{r}
plot.steps.day <- ggplot(impute.activity2, aes(x=daily.step)) + 
  geom_histogram(binwidth = 1000, aes(y=..count.., fill=..count..)) + 
  geom_vline(xintercept=mean.impute, colour="orange", linetype="dotted", size=1) +
  geom_vline(xintercept=median.impute, colour="blue" , linetype="dashed", size=1) +
  labs(title="Number of Steps taken each day (histogram)", y="Frequency", x="Steps per day")
plot.steps.day
```

## 5-Are there differences in activity patterns between weekdays and weekends?
### a) Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
impute.activity$day <- ifelse(weekdays(impute.activity$date) %in% c("Saturday","Sunday"), "weekday", "weekend")
```
##Preparing data for ggplot
```{r}
impute.df <- impute.activity %>% group_by(interval,day) %>% summarise(mean.step=mean(steps))
```
### b)Plot Average steps across weekday/weekend vs 5-min interval Time Series
```{r}
plot.weekday.interval <- ggplot(impute.df, aes(x=interval, y=mean.step, color=day)) + 
  facet_grid(day~.) +
  geom_line() + 
  labs(title="Average Number of Steps Taken vs 5-min Interval on Weekday/Weekend", y="Average Number of Steps", x="5-min Interval Times Series")
plot.weekday.interval
```